Towards interactive model checking\\of biological pathway dynamics

### 2015-01-22 Paper_FORTE15.tex
Reduced conclusions, changed some words and sentences and played with figure/table spacing to start the references on page 15

\newcommand{\ta}{Timed Automaton}
\newcommand{\tas}{Timed Automata}
\newcommand{\nat}{\mathbb{N}}

@@

Biological systems such as regulatory or gene networks can be
seen as a particular type of distributed systems, and for this reason
they can be modelled with the same tools that were developed in the
computer science context. However, tools designed to model distributed
systems often require a computer science background, making their use
less attractive for biologists. ANIMO (Analysis of Networks with
Interactive MOdelling) was built with the purpose to provide biologists
with access to the powerful modelling formalism of Timed Automata
in a user friendly way. Continuous dynamics is handled by discrete approximations.

This brings computational support closer to the biological laboratories, 
where large amounts of data are generated on a daily basis. Thanks to 
computational models, biological data can be analysed in different ways, 
helping biologists to formulate new hypotheses, drive experimental research 
and share knowledge on biological processes.

In this paper we introduce an improved modelling approach that allows us
to considerably increase ANIMO's performances, opening the way for the
analysis of bigger models. Moreover, this improvement makes the introduction
of model checking in ANIMO a realistic feature, allowing for
reduced computation times. The user interface of ANIMO allows to rapidly
build non-trivial models and check them against properties formulated in
a human-readable language, making modelling a powerful
support for biological research.

@@

The current knowledge on signaling pathways (mostly organized in databases such as KEGG~\cite{kegg}
or PhosphoSite~\cite{phosphosite}) suggests that the interactions involved in a cellular response assume more often
the shape of a network than that of a simple chain of signal relays.
Such networks are typically highly connected, involving feedback loops and crosstalk
between multiple pathways, making it difficult for the human
brain to understand their dynamic behavior.
For this reason, a computational support is required when studying non-trivial biological networks.

@@

Previously, ANIMO supported only interactive exploration of network dynamics based on simulation runs. 
Model checking queries could be answered through the UPPAAL tool~\cite{uppaal}, but
the required knowledge of temporal logic together with the usually long response times slowed down the investigation process.
In order to pave the way for a widespread use of model checking on non-trivial models of
signaling networks, we updated ANIMO with a new \tas\ model (presented here). This marks a relevant improvement in terms of
performances with respect to the model previously used in ANIMO. Moreover, consistently with the intents of our tool, we
implemented also a user interface for the definition of model checking queries in ANIMO. This allows a
user to interrogate an ANIMO model without requiring previous experience in temporal logics. These new
features are available in the latest version of ANIMO, which was recently reimplemented as a Cytoscape 3 app~\cite{animo-app-site}:
this lets users profit of the additional analysis features made available by the other apps in the Cytoscape environment.

The rest of the paper is organized as follows: Section~\ref{sec:basics} introduces the basic aspects
of signaling pathways, \tas\ and ANIMO; Section~\ref{sec:animo-new} illustrates a new way of using \tas\ to model
activity-based networks;
Section~\ref{sec:animo-comparison} shows a comparison between the new modeling approach and the
one previously used in ANIMO,
focusing on model analysis performances; Section~\ref{sec:animo-model-checking-ui} describes the
interface additions to make model checking accessible to ANIMO users;
Section~\ref{sec:conclusion} concludes the paper, discussing future work.

@@

The current knowledge on signaling pathways~\cite{kegg,phosphosite} evidences the fact
that signaling interactions are rarely a simple chain of activations as represented in Figure~\ref{fig:small-example-net}.
More often, they assume the shape of a network with multiple feedback loops and crosstalk from different signaling sources.
This complexity is an added difficulty for the study of such networks, reducing the possibilities for the human brain alone
to deduce the dynamic behavior of a signaling network by inspecting its static representation.
For this reason, an efficient computational support is essential when representing and analyzing the behavior
of complex signaling networks.

\subsection{\tas}\label{sec:TA}
\renewcommand{\ta}{TA}
\renewcommand{\tas}{TA}
Timed Automata (\tas) are finite-state automata enriched with real-valued clocks
and synchronization channels. All clocks in a \tas\ system advance with the same rate,
and transitions between the locations of an automaton
depend on conditions on clocks. In particular, a \emph{guard} defines when a transition
may be taken, while an \emph{invariant} is the condition for permanence in a location.
A transition can also allow two automata to \emph{synchronize},
with each participant performing one of two complementary actions (\emph{input} and \emph{output})
on a synchronization channel. A set of clocks may also be reset by a transition, causing them to restart counting from 0.

The models we will present here were implemented using the software tool UPPAAL~\cite{uppaal}, which adds a number of features
to the basic definition of \tas. Some of these extensions include: support for integer variables in addition to clocks,
broadcast synchronization channels (one sender, many receivers), definition of C-like functions to perform more 
operations besides clock resets.
UPPAAL also allows for a special type of locations, named \emph{committed} (marked with a {\sf C} in the graphical representation).
As long as an automaton is in a committed location, time is not allowed to flow. This feature can be used for example to perform immediate
updates to local variables before letting the computation proceed. Examples of the listed features can be found in the
\tas\ model described in Section~\ref{sec:ta-model}.

@@

\begin{figure}[htb]
\begin{center}
   \includegraphics[width=.7\textwidth]{images/ANIMO3_MAPK_model} %mapk_model_egf4}
\end{center}
\caption{The Cytoscape 3 user interface running the new ANIMO app
(model from~\cite{animo-ieee}). The \emph{Network} panel in the center contains the nodes-edges
model of the cross-talking signaling pathways of NGF and EGF, with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first hour of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based. The edge thickness is used to give an idea on which reactions are
occurring more often at the selected time instant.
The series {\sf Erk (EGF) data} in the graph is the experimental
data from~\cite{egf-ngf} for the 100~ng/ml EGF treatment.
\label{fig:cytoscape}}
\end{figure}

@@

The occurrence of a \emph{reaction} modifies by one discrete step the activity level of its target reactant, making it increase or decrease 
depending on whether the reaction is defined, respectively, as activating or inhibiting.
The rate with which a reaction occurs depends on a formula selected by the user. 
Choosing one of three available scenarios allows the user to make the reaction rate depend on
the activity of one or two reactants.
The rate of each reaction can be scaled by modifying the value of one kinetic constant $k$,
possibly using a qualitative measure from a predefined set ({\sf very slow}, {\sf slow},
{\sf medium}, {\sf fast}, {\sf very fast}).
Approximating the actual reaction dynamics with less detailed single-parameter scenarios
leads to a kinetic model that is still precise enough to faithfully represent the behavior of a signaling network.
The approximation allows us to reduce the dependence of a model from often unavailable quantitative
parameters for biochemical reaction kinetics.
For a more precise dissertation on how reaction rates are computed in ANIMO, we recommend~\cite{animo-ieee},
where the previously used \tas\ model is presented. The reader interested in the current
methods for parameter setting in ANIMO can refer to~\cite{animo-syncop}.

@@

\begin{figure*}[!htb]
  \begin{center}
  \subfloat[\label{fig:network-abcd}]{\begin{minipage}[c][\reactantTAheight]{0.18\textwidth}\begin{center}\includegraphics[scale=1.0]{images/network_abcd}\end{center}\end{minipage}} \qquad\qquad
  \subfloat[$R_A$\label{fig:ta-new-model}]{\begin{minipage}[c][\reactantTAheight]{0.6\textwidth}\begin{center}\reactantTA\end{center}\end{minipage}} \\
  \subfloat[$R_1: B \rightarrow A$ \label{fig:ta-old-BA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_B_A}} \qquad\qquad
%   \subfloat[$R_2: C \rightarrow A$ \label{fig:ta-old-CA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_C_A}} \qquad\qquad
  \subfloat[$R_3: D \dashv A$ \label{fig:ta-old-DA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_D_A}}
  \end{center}
  \caption{
{\bf\protect\subref{fig:network-abcd}} A signaling network where one node is influenced by three distinct reactions.
The (virtual) reaction $R_A$ is defined in
the reactant-centered model as the algebraic sum of the reactions influencing $A$.
({\bf\protect\subref*{fig:ta-new-model}\--{}\protect\subref*{fig:ta-old-DA}}) \tas\ templates generated by ANIMO for the reactant-centered~{\bf\protect\subref{fig:ta-new-model}}
 and reaction-centered~({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}) models of the network in~{\bf\protect\subref{fig:network-abcd}}.
The templates have been edited for an easier visual comparison; while the template for $R_2$ ($C \rightarrow A$) is qualitatively identical to the one for $R_1$,
and was left out for the sake of space. Functions {\sf react()}, {\sf can\_react()} and {\sf cant\_react()}
in~{\protect\subref{fig:ta-new-model}} have the same content as the corresponding guards and updates in ({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}),
while function {\sf update()} in~{\protect\subref{fig:ta-new-model}}
replaces the references to the pre-computed tables {\sf tL[]}, {\sf tU[]} with on-the-fly computation (see Sect.~\ref{sec:rates-ta-model}).
% \ta\ template used to represent a net reaction in the reactant-centered model.
\label{fig:ta-models}}
\end{figure*}

As with ODEs, if $r_A$ is positive the activity level of $A$ will increase;
otherwise, $A$ will decrease its activity level.
The absolute value of $r_A$ determines the speed with which such change happens.
The value of the reaction rate is thus translated into a time value to be used as
time bound in the \ta\ representing $R_A$ (see Fig.~\ref{fig:ta-new-model}) by computing
$$T_A = \frac{1}{\mbox{abs}(r_a)}$$
with $\mbox{abs}(r_a)$ the absolute value of $r_a$. In order to represent
a natural variability in reaction timing, we relax the exact value of $T_A$ by
defining bounds of~$\pm~5\%$ (which can be changed by the user):
$$
\begin{array}{lcl}
  \mbox{\sf tL} &=& T_A \times 0.95 \\
  \mbox{\sf tU} &=& T_A \times 1.05
\end{array}
$$
Analogously to the reaction-centered model of~\cite{animo-ieee},
we will call {\sf tL} the \emph{lower time bound} and {\sf tU} the \emph{upper time bound}.
Note that here, differently from what presented in~\cite{animo-ieee}, we do not define
pre-computed tables with all possible values for {\sf tL} and {\sf tU}: this is because
the number of variables for the computation of a reaction rate can be arbitrarily high, and
the size of such tables grows exponentially with the number of variables.
In the example of Figure~\ref{fig:ta-models}, a table to represent the time bounds
for reaction ${R_A}$ would have four dimensions, depending on the activity levels
of reactants $A$, $B$, $C$ and $D$. Assuming that all four reactants are discretized
on 101 levels (activity levels from 0 to 100), each of the two tables {\sf tL} and {\sf tU} would contain $101^4 = 104\,060\,401$
elements. Adding one other reaction $E \rightarrow A$ to the pool would make
this approach exceed the memory limits of most of the currently available computers, requiring about 78 gigabytes of memory
to be stored. For this reason, the reaction rates are computed on the fly (see Section~\ref{sec:rates-ta-model}).

@@

We note that replacing all the existing reactions with the net reactions relative to each
reactant in a model does change the behavior of the model in the sense that continuous oscillations
of activity levels around an equilibrium point are less frequent. However, this does not mean that the
behavior of two models using the two different approaches will diverge: the equilibrium points
are still the same.
Two opposite reactions occurring with the same (or similar) frequency will (nearly) cancel each other out: the
behavior of the single reactions is explicitly represented in the reaction-centered model and gives rise to more oscillations.
On the other hand, in the reactant-centered model, the trajectory of a reactant's activity level
is determined taking into account also the cancelling effect of opposite reactions, so the simulations
will generally contain less oscillations.
From the user's point of view, graphs will look smoother while still exhibiting the same qualitative behavior,
as can be seen in Figure~\ref{fig:comparison-graph-animo}.
The \tas\ templates in Figure~\ref{fig:ta-models} show that the main difference between the two approaches
is indeed due to the choice of lumping all the reactions influencing $A$ into the single
reaction $R_A$.
% Exhibiting less oscillations means generating less events (i.e. changes of state) in the underlying \tas\ model, as in both approaches each change
% in reactant activity level is caused by an event in the \tas\ model. Less events means smaller state spaces, thus lower analysis times.
In the next session we will show that having less oscillations leads in practice to better analysis performances.
\begin{figure}[!htb]
\begin{center}
  \subfloat[Reaction-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reaction_10levels}\label{fig:comparison-graph-animo-reaction}}\qquad\qquad
  \subfloat[Reactant-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reactant_10levels}\label{fig:comparison-graph-animo-reactant}}
\end{center}
  \caption{Comparison of two ERK activity graphs generated from the model based on the case study from~\cite{animo-ieee} with treatment of 50 ng/ml NGF.
  ERK was set to 10 levels of granularity to evidence the oscillations in the graphs.
  \label{fig:comparison-graph-animo}}
\end{figure}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    Model type & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    Reaction-centered & 30.72 & 291\,{}576 \\ %37.14 & 257,928 \\
    \hline
    Reactant-centered & 2.86 & 9\,{}768 \\ %2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-ieee} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution.\label{tab:sim-100}}
\end{table}

@@

\begin{table*}[htbp]
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{Reaction-centered}	   & \multicolumn{2}{c||}{Reactant-centered} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 126.56 & 523\,{}448 & 1.04 & 9\,{}236 & 122 & 57 \\ %127.25 & 351\,{}804 & 1.42 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 159.29 & 436\,{}496 & 1.73 & 11\,{}480 & 92 & 38 \\ %167.64 & 269\,{}932 & 1.93 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 146.09 & 439\,{}484 & 1.04 & 10\,{}384 & 140 & 42 \\ %150.90 & 318\,{}460 & 1.44 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 0.74 & 293\,{}508 & 0.06 & 7\,{}484 & 12 & 39 \\ %2.15 & 216\,{}920 & 0.11 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 581.79 & 448\,{}764 & 6.86 & 16\,{}880 & 85 & 27 \\ %470.56 & 311\,{}568 & 2.22 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 561.01 & 449\,{}248 & 6.42 & 15\,{}852 & 87 & 28 \\ %453.90 & 357\,{}980 & 2.23 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the given queries on the case study from~\cite{animo-ieee}.\label{tab:model-checking}}
\end{table*}

@@

\section{Conclusions and future work}\label{sec:conclusion}
We have presented here how the ANIMO tool 
was improved to provide an interactive approach to model checking.
Thanks to the increased performances of the new reactant-centered modeling approach,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, we ensure that the biologist can access the features
of model checking without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

>>>>

\newcommand{\ta}{Timed Automaton}
\newcommand{\tas}{Timed Automata}
\newcommand{\nat}{\mathbb{N}}
\setlength{\belowcaptionskip}{-10pt}

@@

Biological systems such as regulatory or gene networks can be
seen as a particular type of distributed systems, and for this reason
they can be modeled with the same tools that were developed in the
computer science context. However, tools designed to model distributed
systems often require a computer science background, making their use
less attractive for biologists. ANIMO (Analysis of Networks with
Interactive MOdeling) was built with the aim to provide biologists
with access to the powerful modeling formalism of Timed Automata
in a user friendly way. Continuous dynamics is handled by discrete approximations.

This brings computational support closer to the biological laboratories, 
where large amounts of data are generated on a daily basis. Thanks to 
computational models, biological data can be analyzed in different ways, 
helping biologists to formulate new hypotheses, drive experimental research 
and share knowledge on biological processes.

In this paper we introduce an improved modeling approach that allows us
to considerably increase ANIMO's performances, opening the way for the
analysis of bigger models. Moreover, this improvement makes the introduction
of model checking in ANIMO a realistic feature, allowing for
reduced computation times. The user interface of ANIMO allows to rapidly
build non-trivial models and check them against properties formulated in
a human-readable language, making modeling a powerful
support for biological research.

@@

The current knowledge on signaling pathways (mostly organized in databases such as KEGG~\cite{kegg}
or PhosphoSite~\cite{phosphosite}) suggests that the interactions involved in a cellular response assume more often
the shape of a network than that of a simple chain of signal relays.
Such networks are typically highly connected, involving feedback loops and crosstalk
between multiple pathways, making it difficult for the human
brain to grasp their dynamic behavior.
For this reason, a computational support is required when studying non-trivial biological networks.

@@

Previously, ANIMO supported only interactive exploration of network dynamics based on simulation runs. 
Model checking queries could be answered through the UPPAAL tool~\cite{uppaal}, but
the required knowledge of temporal logic together with the usually long response times slowed down the investigation process.
In order to encourage the use of model checking on non-trivial models of
signaling networks, we updated ANIMO with a new \tas\ model (presented here). This marks a relevant improvement in terms of
performances with respect to the model previously used in ANIMO. Moreover, consistently with the intents of our tool, we
implemented also a user interface for the definition of model checking queries in ANIMO. This allows a
user to interrogate an ANIMO model without requiring previous experience in temporal logics. These new
features are available in the latest version of ANIMO, which was recently reimplemented as a Cytoscape 3 app~\cite{animo-app-site}:
this lets users profit of the additional analysis features made available by the other apps in the Cytoscape environment.

% The rest of the paper is organized as follows:
% Section~\ref{sec:basics} introduces the basic aspects
% of signaling pathways, \tas\ and ANIMO; Section~\ref{sec:animo-new} illustrates a new way of using \tas\ to model
% activity-based networks;
% Section~\ref{sec:animo-comparison} shows a comparison between the new modeling approach and the
% one previously used in ANIMO,
% focusing on model analysis performances; Section~\ref{sec:animo-model-checking-ui} describes the
% interface additions to make model checking accessible to ANIMO users;
% Section~\ref{sec:conclusion} concludes the paper, discussing future work.
The paper continues as follows.
After introducing the basic concepts in Section~\ref{sec:basics}, we illustrate in
Section~\ref{sec:animo-new} a new way of using \tas\ in ANIMO.
In Section~\ref{sec:animo-comparison} we present a comparison between the new modeling approach and the
one previously used in ANIMO,
focusing on model analysis performances. In Section~\ref{sec:animo-model-checking-ui} we describe
how model checking was made accessible to ANIMO users.
We conclude the paper with Section~\ref{sec:conclusion}, discussing future work.

@@

The current knowledge on signaling pathways~\cite{kegg,phosphosite} evidences the fact
that signaling interactions are rarely a simple chain of activations as represented in Figure~\ref{fig:small-example-net}.
More often, they assume the shape of a network with multiple feedback loops and crosstalk from different signaling sources.
This complexity is an added difficulty for the study of such networks, reducing the possibilities
%for the human brain alone
to deduce the dynamic behavior of a signaling network by inspecting its static representation.
For this reason, an efficient computational support is essential when representing and analyzing the behavior
of complex signaling networks.

\subsection{\tas}\label{sec:TA}
\renewcommand{\ta}{TA}
\renewcommand{\tas}{TA}
Timed Automata (\tas) are finite-state automata enriched with real-valued clocks
and synchronization channels. All clocks in a \tas\ system advance with the same rate,
and transitions between the locations of an automaton
depend on conditions on clocks. In particular, a \emph{guard} defines when a transition
may be taken, while an \emph{invariant} is the condition for permanence in a location.
A transition can also allow two automata to \emph{synchronize},
with each participant performing one of two complementary actions (\emph{input} and \emph{output})
on a synchronization channel. A set of clocks may also be reset by a transition, causing them to restart from 0.

The models we will present here were implemented using the software tool UPPAAL~\cite{uppaal}, which adds a number of features
to the basic definition of \tas. Some of these extensions include: support for integer variables in addition to clocks,
broadcast synchronization channels (one sender, many receivers), definition of C-like functions to perform more 
operations besides clock resets.
UPPAAL also allows for a special type of locations, named \emph{committed} (marked with a {\sf C} in the graphical representation).
As long as an automaton is in a committed location, time is not allowed to flow. This feature can be used for example to perform immediate
updates to local variables before letting the computation proceed. Examples of the listed features are found in the
\tas\ model in Section~\ref{sec:ta-model}.

@@

\begin{figure}[htb]
\begin{center}
   \includegraphics[width=.7\textwidth]{images/ANIMO3_MAPK_model} %mapk_model_egf4}
\end{center}\vspace{-.5cm}
\caption{The Cytoscape 3 user interface running the new ANIMO app
(model from~\cite{animo-ieee}). The \emph{Network} panel in the center contains the nodes-edges
model of the cross-talking signaling pathways of NGF and EGF, with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first hour of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based. The edge thickness is used to give an idea on which reactions are
occurring more often at the selected time instant.
The series {\sf Erk (EGF) data} in the graph is the experimental
data from~\cite{egf-ngf} for the 100~ng/ml EGF treatment.
\label{fig:cytoscape}}
\vspace{-.5cm}
\end{figure}

@@

The occurrence of a \emph{reaction} modifies by one discrete step the activity level of its target reactant, making it increase or decrease 
depending on whether the reaction is defined, respectively, as activating or inhibiting.
The rate with which a reaction occurs depends on a formula selected by the user. 
Choosing one of three available scenarios allows the user to make the reaction rate depend on
the activity of one or two reactants.
The rate of each reaction can be scaled by modifying the value of one kinetic constant $k$,
possibly using a qualitative measure from a predefined set ({\sf very slow}, {\sf slow},
{\sf medium}, {\sf fast}, {\sf very fast}).
% Approximating the actual reaction dynamics with less detailed single-parameter scenarios
% leads to a kinetic model that is still precise enough to faithfully represent the behavior of a signaling network.
The approximation allows us to reduce the dependence of a model from often unavailable quantitative
parameters for biochemical reaction kinetics, while keeping a precision level that is still precise enough to be useful.
For a more precise dissertation on how reaction rates are computed in ANIMO, we recommend~\cite{animo-ieee},
where the previously used \tas\ model is presented. The reader interested in the current
methods for parameter setting in ANIMO can refer to~\cite{animo-syncop}.

@@

\begin{figure}[thb]
  \begin{center}
  \subfloat[\label{fig:network-abcd}]{\begin{minipage}[c][\reactantTAheight]{0.26\textwidth}\begin{center}\includegraphics[scale=1.0]{images/network_abcd}\end{center}\end{minipage}} \qquad\qquad\quad
  \subfloat[$R_A$\label{fig:ta-new-model}]{\begin{minipage}[c][\reactantTAheight]{0.6\textwidth}\begin{center}\reactantTA\end{center}\end{minipage}} \\
  \subfloat[$R_1: B \rightarrow A$ \label{fig:ta-old-BA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_B_A}} \qquad\qquad
%   \subfloat[$R_2: C \rightarrow A$ \label{fig:ta-old-CA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_C_A}} \qquad\qquad
  \subfloat[$R_3: D \dashv A$ \label{fig:ta-old-DA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_D_A}}
  \end{center}\vspace{-.5cm}
  \caption{
{\bf\protect\subref{fig:network-abcd}} A signaling network where one node is influenced by three distinct reactions.
The (virtual) reaction $R_A$ is defined in
the reactant-centered model as the algebraic sum of the reactions influencing $A$.
({\bf\protect\subref*{fig:ta-new-model}\--{}\protect\subref*{fig:ta-old-DA}}) \tas\ templates generated by ANIMO for the reactant-centered~{\bf\protect\subref{fig:ta-new-model}}
 and reaction-centered~({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}) models of the network in~{\bf\protect\subref{fig:network-abcd}}.
The templates have been edited for an easier visual comparison; while the template for $R_2$ ($C \rightarrow A$) is qualitatively identical to the one for $R_1$,
and was left out for the sake of space. Functions {\sf can\_react()} and {\sf cant\_react()}
in~{\protect\subref{fig:ta-new-model}} have the same content as the corresponding guards in ({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}),
while function {\sf update()} in~{\protect\subref{fig:ta-new-model}}
replaces the references to the pre-computed tables {\sf tL[]}, {\sf tU[]} with on-the-fly computation (see Sect.~\ref{sec:rates-ta-model}).
% \ta\ template used to represent a net reaction in the reactant-centered model.
\label{fig:ta-models}}
\end{figure}

As with ODEs, if $r_A$ is positive the activity level of $A$ will increase;
otherwise, $A$ will decrease its activity level.
The absolute value of $r_A$ determines the speed with which such change happens.
The value of the reaction rate is thus translated into a time value to be used as
time bound in the \ta\ representing $R_A$ (see Fig.~\ref{fig:ta-new-model}) by computing
$$T_A = \frac{1}{\mbox{abs}(r_a)}$$
with $\mbox{abs}(r_a)$ the absolute value of $r_a$. In order to represent
a natural variability in reaction timing, we relax the exact value of $T_A$ by
defining bounds of~$\pm~5\%$ (which can be changed by the user):
$$
\begin{array}{lcl}
  \mbox{\sf tL} &=& T_A \times 0.95 \\
  \mbox{\sf tU} &=& T_A \times 1.05
\end{array}
$$
Analogously to the reaction-centered model of~\cite{animo-ieee},
we will call {\sf tL} the \emph{lower time bound} and {\sf tU} the \emph{upper time bound}.
Note that here, differently from what presented in~\cite{animo-ieee}, we do not define
pre-computed tables with all possible values for {\sf tL} and {\sf tU}: this is because
the number of variables for the computation of a reaction rate can be arbitrarily high, and
the size of such tables grows exponentially with the number of variables.
In the example of Figure~\ref{fig:ta-models}, a table to represent the time bounds
for reaction ${R_A}$ would have four dimensions, depending on the activity levels
of reactants $A$, $B$, $C$ and $D$. Assuming that all four reactants are discretized
on 101 levels (activity levels from 0 to 100), each of the two tables {\sf tL} and {\sf tU} would contain $101^4 = 104\,060\,401$
elements. Adding one more reaction $E \rightarrow A$ to the pool would make
this approach exceed the memory limits of most of the currently available computers
% , requiring about 78 gigabytes of memory to be stored.
For this reason, the reaction rates are computed on the fly (see Section~\ref{sec:rates-ta-model}).

@@

We note that replacing all the existing reactions with the net reactions relative to each
reactant in a model does change the behavior of the model in the sense that continuous oscillations
of activity levels around an equilibrium point are less frequent. However, this does not mean that the
behavior of two models using the two different approaches will diverge: the equilibrium points
are still the same.
Two opposite reactions occurring with the same (or similar) frequency will (nearly) cancel each other out: the
behavior of the single reactions is explicitly represented in the reaction-centered model and gives rise to more oscillations.
On the other hand, in the reactant-centered model, the trajectory of a reactant's activity level
is determined taking into account also the canceling effect of opposite reactions, so the simulations
will generally contain less oscillations.
From the user's point of view, graphs will look smoother while still exhibiting the same qualitative behavior,
as can be seen in Figure~\ref{fig:comparison-graph-animo}.
The \tas\ templates in Figure~\ref{fig:ta-models} show that the main difference between the two approaches
is indeed due to the choice of lumping all the reactions influencing $A$ into the single
reaction $R_A$.
% Exhibiting less oscillations means generating less events (i.e. changes of state) in the underlying \tas\ model, as in both approaches each change
% in reactant activity level is caused by an event in the \tas\ model. Less events means smaller state spaces, thus lower analysis times.
In the next session we will show that having less oscillations leads in practice to better analysis performances.
\begin{figure}[!htb]
\begin{center}
  \subfloat[Reaction-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reaction_10levels}\label{fig:comparison-graph-animo-reaction}}\qquad\qquad
  \subfloat[Reactant-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reactant_10levels}\label{fig:comparison-graph-animo-reactant}}
\end{center}\vspace{-.5cm}
  \caption{Comparison of two ERK activity graphs generated from the model based on the case study from~\cite{animo-ieee} with treatment of 50 ng/ml NGF.
  ERK was set to 10 levels of granularity to evidence the oscillations in the graphs.
  \label{fig:comparison-graph-animo}}
\end{figure}

@@

\begin{table}
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    Model type & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    Reaction-centered & 30.72 & 291\,{}576 \\ %37.14 & 257,928 \\
    \hline
    Reactant-centered & 2.86 & 9\,{}768 \\ %2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-ieee} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution.\label{tab:sim-100}}
\vspace{-1cm}
\end{table}

@@

\begin{table}[htbp]
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{Reaction-centered}	   & \multicolumn{2}{c||}{Reactant-centered} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 126.56 & 523\,{}448 & 1.04 & 9\,{}236 & 122 & 57 \\ %127.25 & 351\,{}804 & 1.42 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 159.29 & 436\,{}496 & 1.73 & 11\,{}480 & 92 & 38 \\ %167.64 & 269\,{}932 & 1.93 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 146.09 & 439\,{}484 & 1.04 & 10\,{}384 & 140 & 42 \\ %150.90 & 318\,{}460 & 1.44 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 0.74 & 293\,{}508 & 0.06 & 7\,{}484 & 12 & 39 \\ %2.15 & 216\,{}920 & 0.11 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 581.79 & 448\,{}764 & 6.86 & 16\,{}880 & 85 & 27 \\ %470.56 & 311\,{}568 & 2.22 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 561.01 & 449\,{}248 & 6.42 & 15\,{}852 & 87 & 28 \\ %453.90 & 357\,{}980 & 2.23 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the given queries on the case study from~\cite{animo-ieee}.\label{tab:model-checking}}
\vspace{-.5cm}
\end{table}

@@

\section{Conclusions and future work}\label{sec:conclusion}
We have presented here how the ANIMO tool 
was improved to provide a more interactive modeling process.
Thanks to the increased performances of the new reactant-centered modeling approach,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, the features
of model checking are made accessible without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

In order to enforce the concept of user interaction as a primary focus of the tool, we plan to extend
ANIMO by increasing the support for parameter sensitivity analysis and parameter fitting,
as a follow-up to what was presented in~\cite{animo-syncop}.
% We already allow for only one numeric parameter per reaction
% and a ``brute force'' parameter sweep is already available, but a more automatic way of adjusting such
% parameters to fit experimental data will allow the user to concentrate more on defining a sensible topology for a pathway.
% Such an automatic search could be limited by specific bounds based on previous knowledge,
% or on a trade-off with performances.
Moreover, inspired by works on automata learning~\cite{test-based-modelling}, we plan to add also the possibility
to automatically derive a network topology based on experimental data and 
previous knowledge.
